Measuring and optimizing machine learning interpretability

### 2019-03-18 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
pdf; continue

---
title: "Measuring and optimizing machine learning interpretability"
author: "Christoph Molnar"
date: "2019-03-21"
output: 
  ioslides_presentation:
    widescreen: false
    smaller: false
css: ../styles.css
---

@@

```{r, echo = FALSE}
knitr::include_graphics("../images/black-box.png")
```

## Solution 1: Interpretable Model

```{r, echo=FALSE}
knitr::include_graphics("../images/white-box.png")
```

## But: Peformance low and how to compare interpretability?

```{r}
knitr::include_graphics("../images/simple-models.png")
```

## Solution 2: Post-Hoc Analysis

@@

## But: Verbose and unreliable {.center data-background=../images/pdp-multiple.png}

```{r, eval=FALSE}
library('iml')
dat = data.frame(matrix(rnorm(1800), ncol = 180))
pred.fun = function(newdata, object){
	data.frame(y = sin(rowSums(newdata)) + rnorm(nrow(newdata), sd = 0.1))
}
pred = Predictor$new(data = dat, predict.fun = pred.fun)
fes = FeatureEffects$new(pred)
fes$plot(ncol = 20)
```
```{r eval=FALSE}
knitr::include_graphics("../images/pdp-multiple.png")
```

## Solution 3: Compare models with interpretability / complexity measures

TODO: Image with 3 black boxes and numbers attached to it

## Desiderata

- Few features (Sparsity)
- Few interactions (Additivity)
- Simple relationships (Simplicity)

## Measure: Number of features heuristic

TODO: Visualisation of algorithm

@@

TODO: Formula with decomposition

@@

```{r}
knitr::include_graphics("../images/ale-intuition-3.jpg")
```

## Functional Decomposition with ALE

TODO: copy paste new decomposition

@@

TODO: PRL formula with R2 and SSQ and so on

## Measure: Complexity

TODO: AleNumApprox plot

## Measure: Complexity cont.

TODO: Multiple AleNumApprox plots with averaging formula

@@

```{r child = "../chunks/book.Rmd"}
```

>>>>

---
title: "Measuring and optimizing machine learning interpretability"
author: "Christoph Molnar, Giuseppe Casalicchio, Bernd Bischl"
date: "2019-03-21"
output: 
    beamer_presentation
---

@@

```{r, echo = FALSE}
knitr::include_graphics("../images/black-box.png")
```
## Problem: Tradeoff interpretability and performance
Use only interpretable models (e.g. linear model)! $\Rightarrow$ Pay with loss in performance.

```{r, echo = FALSE}
knitr::include_graphics("../images/white-box.png")
```
## Problem: Tradeoff interpretability and performance
Just use post-hoc interpretation methods (e.g. feature importance)! $\Rightarrow$ Works better for less complex models.

@@

## What's interpretability anyway?

TODO of fuzzy animal which dodges touch of hand
Or use image of cat that doges hand

## We propose

TODO: Replace text with image: Black box in the middle, three properties attached to it

Measure model complexity in a model-agnostic way: number of features, interaction strength, main effect complexity

$\Rightarrow$ Allows model comparision

$\Rightarrow$ Allows direct optimization for interpretability

$\Rightarrow$ Makes claims of interpretability more explicit ("Model A uses less features than B and has less interactions)

@@

$$f(x) = \overbrace{f_0}^\text{Intercept} + \overbrace{\sum_{j=1}^p f_j(x_j)}^\text{1st order effects} + \overbrace{\sum_{j\neq k}^p f_{jk}(x_j, x_k)}^\text{2nd order effects} + \ldots + \overbrace{f_{1,\ldots,p}(x_1, \ldots, x_p)}^\text{p-th order effect}
$$

$$f(x)  = \overbrace{f_0}^\text{Intercept} + \overbrace{\sum_{j=1}^p f_j(x_j)}^\text{1st order effects} + \overbrace{\sum_{S \subseteq \{1,\ldots,p\},|S| \geq 2} f_{S}(x_S)}^{\text{Higher order effects}}
$$

$$f(x)  = \underbrace{f_0 + \sum_{j=1}^p \overbrace{f_j(x_j)}^\text{How complex?} + \overbrace{IA(x)}^{\text{How much interaction?}}}_{\text{How many feature used?}}
$$

@@

TODO: Use all plots

```{r}
knitr::include_graphics("../images/ale-intuition-3.jpg")
```
## Measure: Main effect complexity

TODO: Plot of single ALE plot with complexity (from paper)

Plus formula how to average those for all plots

@@

Measure main effect model with proportion of error explained (pre):

$$\text{Interaction Strength} =  \frac{\mathbb{E}(L(\hat{f}, f_{ALE}))}{\mathbb{E}(L(\hat{f}, c))}$$

@@
